---
title: "Logistic regression on credit data"
author: "Jarod Barnhart"
date: "5/30/2022"
output:
  html_document: default
---
This project is based around creating a logistic regression model to predict
whether or not a client will fulfill their credit contract based on many 
different predictors. The prediction is a binary outcome response variable. The
data is also in German, so some translation is required.
``` {r, include=FALSE}
library('rmarkdown')
library('glmnet')
library('tidyverse')
library('ROSE')
library('caTools')
library('leaps')
library('caret')
```

``` {r, include=FALSE}
#Reading in data
kredit<-read.table('data/SouthGermanCredit.asc',header=T)
names<-c('status','duration','credit_history','purpose','amount','savings',
         'employment_duration','installment_rate','personal_status_sex',
         'other_debtors','present_residence','property','age','other_installment_plans',
         'housing','number_credits','job','people_liable','telephone',
         'foreign_worker','credit_risk')
colnames(kredit)<-names

#Coercing categorical variables to factors
categorical<-c('status','credit_history','purpose','savings','personal_status_sex',
               'other_debtors','other_installment_plans','housing','telephone',
               'foreign_worker','credit_risk')

kredit[categorical] <- lapply(kredit[categorical],factor)
```

``` {r echo=FALSE}
#Splitting into training and testing data
smp_size <- floor(0.75*nrow(kredit))
set.seed(1337)
train_ind<-sample(seq_len(nrow(kredit)), size = smp_size)
kredit_train <- kredit[train_ind, ]
kredit_test <- kredit[-train_ind, ]
```
After reading in our data, and splitting into training and testing, we make an 
initial logistic regression model.
```{r}
# Logistic regression
kredit_logit <- glm(credit_risk~status+duration+credit_history+purpose+amount+
                      savings+employment_duration+installment_rate+personal_status_sex+
                      other_debtors+present_residence+property+age+other_installment_plans+
                      housing+number_credits+job+people_liable+telephone+foreign_worker,
                    data=kredit_train,family=binomial)
summary(kredit_logit)
```
There are quite a few variables which are factors, and many levels are 
insignificant. From here we use sequential replacement variable selection,
maxing out the number of variables at 11
``` {r, results="hide"}
#Sequential replacement variable selection with max 11 variables
kredit_models<-regsubsets(credit_risk~status+duration+credit_history+purpose+amount+
                            savings+employment_duration+installment_rate+personal_status_sex+
                            other_debtors+present_residence+property+age+other_installment_plans+
                            housing+number_credits+job+people_liable+telephone+foreign_worker,
                          data = kredit_train, nvmax = 11, method = "seqrep")
summary(kredit_models)
```
We can reduce our model down to a few significant variables.
```{r}
kredit_reduced<-glm(credit_risk~status+credit_history+purpose+amount+savings+
                      installment_rate+telephone,
                    data=kredit_train,family=binomial)
summary(kredit_reduced)
```
Let us test this reduced logistic model.
```{r}
kredit_reduced_test<-glm(credit_risk~status+credit_history+purpose+amount+savings+
                           installment_rate+telephone,
                         data=kredit_test,family=binomial)
summary(kredit_reduced_test)
```
Very few of the variables which were significant in the training model are 
significant on the testing data. Lets construct an ROC curve to analyze
performance.
```{r}
### c.) ROC Curve
prob_train <- predict(kredit_logit, type = "response")
pred_train<- ifelse(prob_train > 0.5, "1", "0")
prob_test <- predict(kredit_logit, kredit_test, type = "response")
pred_test<- ifelse(prob_test > 0.5, "1", "0")
probabilities_train <- predict(kredit_reduced, type = "response")
predicted.classes_train<- ifelse(probabilities_train > 0.5, "1", "0")
probabilities_test <- predict(kredit_reduced, kredit_test, type = "response")
predicted.classes_test<- ifelse(probabilities_test > 0.5, "1", "0")

roc.curve(kredit_train$credit_risk, predicted.classes_train) #AUC=0.673
roc.curve(kredit_test$credit_risk, predicted.classes_test) #AUC=0.640
roc.curve(kredit_train$credit_risk, pred_train) #AUC=0.712
roc.curve(kredit_test$credit_risk, pred_test) #AUC0.691
```
